Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-22 workspace/heyuan.Rmd
Add files via upload

## POMP 1

```{r}
install.packages("pomp")
library(pomp)
```

To create a POMP for CPUE, I need to define the state process that reflects the underlying dynamics of the species population and the observation process that relates the state to the CPUE measurements.

>>>>

## POMP 3

Y(t): bird count, $y_t = NB(mean = \rho B_t, \sigma)$, $B_t$ is bird population size at time t
X(t): unobserved state process, red foxes, Bird, Peak Rodent Year, temperature, precipitation, NAO index

Reference: Lotka-Volterra predator-prey model for the state process: $$X_{t+dt} = \beta X_t - \mu X_t$$

natural born - number of birds eaten by foxes: 
$$B_{t+dt} = \alpha B_t - \beta(R_t) B_{t}F_{t}$$

where, $$\beta(R_t) = \beta if R(t)=0$$ or 
$$ \beta(R_t) = \gamma \beta, \gamma > 1$$ , rodent number affect the rate of fox eating birds, 
$\beta(R_t)$ can be re-written as $$\beta(R_t) =  \gamma^R \beta$$, where R = 0 or 1

Growth Rate - Death Rate of Foxes: 
$$ F_{t+dt} = \delta(t) F_t B^*_t - \theta(t) F_t$$

where, $$\delta(t) = \delta_1 if R(t) = 0, or, \phi \delta_1$$ , $\phi<1$ and whether it is in rodent peak year affect the number of birds and the number of foxes.

$\delta(t)$ can be re-written as $$\delta(t) = \phi^R \delta_1$$, where R = 0 or 1

$\theta(t)$ could include the temperature, NAO about how does it affect the death rate of Foxes.

first, let dt=1 year
then, dt = less than 1 year, maybe 1 month, bird keeps as 100, foxes=10, 15, 17, 15 for each dt until reaching 1 year. 
then include bird in X(t), lecture 16

another way: bird is not in X(t)

$$B_{t+dt} = \alpha B^*_{t+dt} - \beta(R_t) B_{t} F_{t}$$

just do the fox equation, no bird equation

```{r}
# compute the average temperature and preciptation. 
data_frame <- df[,c(1,3,6,7,8:17)]
data_frame$Temp <- rowMeans(data_frame[,c(5:10)], na.rm = TRUE) 
data_frame$Precip <- rowMeans(data_frame[,c(11:14)], na.rm = TRUE) 
data_frame <- data_frame[,c(1,2,3,4,15,16)]
names(data_frame) <- c('year', 'CUPE', 'R', 'NAO', 'Temp', 'Precip')
data_frame$R <- ifelse(data_frame$R == "yes", 1, 0)
data_frame <- transform(data_frame, year = as.numeric(year))  # Ensure 'year' is numeric

Euler Approximation of the predator population at time $t+ \delta t$...

$$F_{t+\delta t} = F_t + \delta(t) F_t B_t \delta t - \theta F_t \delta t$$
$$F_{t+\delta t} = F_t + (\phi^R \delta_1 F_t B_t - \theta F_t) \delta t$$

$\delta(t) F_t B_t \delta t$ is the expected number of new predators arising from the Poisson process with rate $\delta(t) F_t B_t$;
$- \theta F_t \delta t$ is the deterministic decrease in predators due to natural death over the time step $\delta t$

$$B_{t+ \delta t} = B_t + \alpha B_t \delta t - \beta(R_t) B_t F_t\delta t$$
$$B_{t+ \delta t} = B_t + (\alpha B_t - \gamma^R \beta B_t F_t) \delta t$$

```{r, warning=FALSE}
meas <- dplyr::select(data_frame, year, reports=CUPE, R)

sir_step<- Csnippet("
  double rateBin, rateBout, rateFin, rateFout;

  rateBin = alpha * B * dt;
  rateBout = pow(gamma, R) * beta_1 * B * Fox * dt;
  rateFin = pow(psi, R) * delta_1 * Fox * B * dt;
  rateFout = theta * Fox * dt;

  // double dF = rbinom(Fox, 1 - exp(-rateF));
  double dFin = rbinom(Fox, 1 - exp(-rateFin));
  double dFout = rbinom(Fox, 1 - exp(-rateFout));

  double dBin = rbinom(B, 1 - exp(-rateBin));
  double dBout = rbinom(B, 1 - exp(-rateBout));

  Fox += dFin - dFout;  
  B += dBin - dBout;
  B = fmax(B, 0.0); 
  Fox = fmax(Fox, 0.0);
  // Rprintf(\"B = %f\\n\\n \", B);

")

# dmeasure: Define the measurement model
sir_dmeas <- Csnippet("
  if (give_log) {
    if (B == 0 & reports != 0) {
      lik = -100000;
    } else {
      // Rprintf(\"rho * B = %f\\n \", rho * B);
      // Rprintf(\"reports = %f\\n\\n \", reports);
      lik = dnbinom_mu(reports, k, rho * B, give_log);
    }
  } else {
    lik = dnbinom_mu(reports, k, rho * B, give_log);
  }
")

# rmeasure: 
sir_rmeas <- Csnippet("
  reports = rnbinom_mu(k, rho * B);
")
## size =  number of successes until the experiment is stopped; success rate / hunting rate:

# initialize:
sir_rinit <- Csnippet("
  Fox = nearbyint(F_0);
  B = nearbyint(B_0);
")

# Define parameters and initial states
params <- c(alpha = 0.5, beta_1 = 0.0001, gamma = 1, delta_1 = 0.01, psi = 3, theta = 0.1, rho = 0.01, k = 20, F_0 = 90, B_0 = 1000)

my_covars <- covariate_table(meas |> dplyr::select(R, year) |> as.data.frame(), times = 'year', order = 'constant')

# create a pomp object
pomp(data = meas |> dplyr::select(-R) |> as.data.frame(),
     rprocess=euler(sir_step,delta.t=1/12), times='year', t0=1871,
    rinit=sir_rinit,
    rmeasure=sir_rmeas,
    dmeasure=sir_dmeas,
    statenames=c("Fox","B"),
    paramnames=c("alpha", "gamma", "delta_1", 'beta_1', "psi", "theta", "rho" ,'k', 'F_0', 'B_0'),
    covar = my_covars,
    partrans = parameter_trans(log = c('alpha', 'gamma', 'delta_1', 'beta_1', 'psi', 'theta', 'rho', 'k', 'F_0', 'B_0'))
  ) -> measSIR

measSIR |>
  pfilter(Np=5, params = params) -> pf

mif2(
  measSIR, 
  Nmif = 10, 
  Np = 2000,
  cooling.fraction.50 = 0.5, 
  rw.sd = rw_sd(
    alpha = 0.02, beta_1 = 0.02, gamma = 0.00, delta_1 = 0.02, psi = 0.02, rho = 0.02, k = 0.02, F_0 = ivp(0.1), B_0 = ivp(0.1), theta = 0.02
  ),
  params = params
) -> mif_out

```{r}
statenames <- c('logF', 'logB')
paramnames <- c('a', 'b', 'c', 'gamma', 'alpha', 'Beta', 'sigmaF', 'sigmaB',
                'logF_0', 'logB_0', 'logRho', 'sigma_obs')
obs_names <- c('logCPUE')
# logF_0 = 0 normalizing to assume unit initial population
# sigmaF roughly the noise in a year
param_guess <- c(
  logF_0 = 0,
  logB_0 = 0,
  a = 1,
  b = 2,
  c = 1,
  alpha = 1,
  Beta = 1,
  gamma = 0.5,
  sigmaF = 0.1,
  sigmaB = 0.1,
  logRho = 3,
  sigma_obs = 0.1
)
Csnippet('
  double dwB, dwF;
  dwF = rgammawn(sigmaF, dt);
  dwB = rgammawn(sigmaB, dt);
  logF += (b*R + c*exp(logB)*(1-gamma*R)-a)*dwF;
  logB += (alpha - Beta*exp(logF)*(1-gamma*R))*dwB;
') -> rproc
Csnippet('
  logF = logF_0;
  logB = logB_0;
') -> rinit
Csnippet('
  log.CPUE = rnorm(logB + logrho, signma_obs);
') -> rmeas
Csnippet('
  lik = dnorm(logCPUE, logB + logrho, sigma_obs, give_log);
') -> dmeas

my_covars <- covariate_table(meas |> dplyr::select(R, year) |> as.data.frame(), times = 'year', order = 'constant')

pomp(data = data_frame |> dplyr::select(-R),rprocess=euler(rproc,delta.t=1/12),
     times = 'year', t0 = 1871,
    rinit=rinit,
    # rmeasure=rmeas,
    # dmeasure=dmeas,
    statenames=statenames,
    paramnames=paramnames,
    covar = my_covars
  ) -> measSIR
